---
title: "Comparing curated TE libraries"
date: 2025-08-07
execute:
  warning: false
format:
  html:
    self-contained: true
    fig-height: 12
    fig-width: 12
    fig-format: png
    fig-dpi: 300
---


# Comparing results of different TEammo curators

## Run library compare
## 0. Prerequisites
- Make conda env for blast
```{bash}
module load conda
conda create -n blast -c conda-forge -c bioconda blast
```

## 1. Run blast library comparison between your two libraries
```{bash}

source bash_vars_Dpla

conda activate blast

scripts/lib_compare2.sh \
${lib1} \
${curator1} \
${lib2} \
${curator2} \
${species} \
${strain} \
${auto_mchelper_lib}

```


## Load and overall comparisons

```{r}
#| output: false
source("scripts/mccompare_functions.R")

library(plotly)
library(dplyr)
library(ggplot2)
library(Biostrings)
library(tidyr)
library(seqinr)
library(patchwork)
library(forcats)
library(gt)

# Color schemes

palette_seq_match <- c(
  "Missing from lib1" = "black",                     
  "Missing from lib2" = "grey50",                  
  "Present, <70" = RColorBrewer::brewer.pal(2, "YlGn")[2],
  "Present, 70" = RColorBrewer::brewer.pal(3, "YlGn")[3],
  "Present, 80" = RColorBrewer::brewer.pal(4, "YlGn")[4],
  "Perfect, 95-100" = RColorBrewer::brewer.pal(5, "YlGn")[5]
)

# Load the TE classifications
teClassification <- read.table("orozco_classification-2024_mchelper.csv", sep = ";", header = TRUE)
teClassification <- teClassification %>%
  mutate(AppName = paste(Class, Order, Superfamily, sep ="/") %>% sub("/$", "", .) %>% sub("/$", "", .))
teClassification <- rbind(
  teClassification
  , c("", "Unclassified", "", "Unclassified")
)

```

# Load BLAST results and plot comparison

- Here we plot the outputs from `scripts/lib_compare2.sh`

```{r}

Dpla <- LoadBlastComparison2(
  "output/lib_compare/lib_compare_MCHelperAuto_vs_D.planitibia-NA2-Dean.blast.out",
  "output/lib_compare/lib_compare_MCHelperAuto_vs_D.planitibia-NA2-Marta.blast.out",
  "D.planitibia",
  "NA2",
  "Dean_vs_Marta",
)

PlotBlastBarMatches2(Dpla)


```

# Extract and examine TEs not shared between libraries
## Rerun teaid for the missing TEs

```{bash}
source bash_vars_Dpla

bash scripts/rerun_teaid.sh ${auto_mchelper_lib} ${species} ${strain} ${lib1} ${curator1} ${lib2} ${curator2} ${busco_lib} ${genome_path}

```

